\vskip-0.6in Trade Agreements in the Shadow of Lobbying\protect\footnote{Formerly circulated under the title ``Trade Agreements, Lobbying and Separation of Powers''}

### 2016-06-13 draft.tex
finish revising model section

Figure~\ref{fig:ext} illustrates the timing of the game from the perspective of the home country.\footnote{This extensive form is for the case where the foreign legislature does not break the trade agreement.} First, the executives set trade policy cooperatively within the context of an international agreement by choosing the trade agreement tariffs $\bta = \left(\tau^a,\tau^{*a}\right)$. Then the lobbies attempt to persuade the legislators in their respective countries to break the trade agreement by choice of lobbying effort $e_b$, which can take on any non-negative value.

@@

\subsection{Preferences}
\label{sec:pref}
As the economy is fully separable and the economic and political structures are symmetric, I focus throughout the paper on the home country and the $X$-sector. The details are analogous for $Y$ and foreign.\footnote{These are the relevant sectors for analysis because they are the sectors in which there are lobbies.}

I specify the welfare function of the home legislature as
\begin{equation}
  W_{\mathit{ML}} = \mathit{CS}_X(\tau) + \ga(e,\ve) \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
  \label{eq:ml}
\end{equation}
where $\mathit{CS}$ is consumer surplus, $\mathit{PS}$ is producer surplus, $\ga(e,\ta)$ is the weight placed on producer surplus (profits) in the import-competing industry, $e$ is lobbying effort, $\ta$ is a random variable, and $\mathit{TR}$ is tariff revenue.\footnote{Labor income $l$ could also be included in both median legislator and executive welfare. I omit it because its inclusion alters none of the results and only serves to complicate the exposition.} I model the decisions of the legislature as being taken by a median legislator. Notice that the weight the median legislator places on the profits of the import-competing industry, $\ga(e,\ve)$, is affected by the level of lobbying effort and a random variable $\ta$. I make the following assumptions on $\ga(e,\ta)$:

@@

Assumption~\ref{as:ga_c} formalizes the intuition that the legislature favors the import-competing industry more the higher is its lobbying effort, but that there are diminishing returns to lobbying activity.\footnote{The diminishing returns here take the form of declining increments to the lobby's influence as effort increases; in Ethier (2012), the returns to lobbying decline as the level of protection increases.} It allows the political uncertainty inherent in $\ta$ to be correlated in interesting ways with the level of $e$, ruling out only (i) that higher effort levels make lower $\ga$-weights more likely and (ii) that effort and uncertainty are correlated in such a way that increasing lobbying effort changes the structure of uncertainty so that higher $\ga$-weights become more likely at an accelerating pace. Assumption~\ref{as:ga_ta} \textbf{simply provides for an intuitive labelling so that} larger realization of the uncertainty variable $\ta$ increase the value of the political economy weight.

Given its expectations and the legislature's preferences, the home lobby chooses its lobbying effort ($e_b$ to influence the break/ratification decision, which for simplicity, I will refer to henceforth simply as ``break,'' and $e_n$ to influence the trade war tariff) to maximize the welfare function:
\begin{equation}
  \expect \left[U_L \right] = \Pr\left[ \text{TradeWar}(\bta,e_b,e_n) \right] \left[ \pi(\tau^{\mathit{n}}) - e_n \right] + \Pr\left[ \text{TradeAgreement}(\bta,e_b,e_n) \right] \pi(\tau^a) - e_b
  \label{eq:lv}
\end{equation}
where $\pi(\cdot)$ is the current-period profit and $\tau^a \ (\tau^\mathit{n})$ is the home country's tariff on the import good under a trade agreement (war). I use the convention throughout of representing a vector of tariffs for both countries $(\tau,\tau^*)$ as a single bold $\bm{\tau}$.

In the first stage of the game, the executives choose the trade agreement tariffs $\bta=\left(\tau^a,\tau^{*a} \right)$ via a negotiating process that I assume to be efficient. In this symmetric environment, it therefore maximizes the expected joint payoffs:
\begin{multline}
  \expect \left[W_E(\bta) \right] + \expect \left[ W_E^{*}(\bta) \right] = \Pr\left[ \text{TradeWar}(\bta) \right] \left[W_E(\bm{\tau}^{\mathit{n}}) + W_E^*(\bm{\tau}^{\mathit{n}}) \right] + \\ \Pr\left[ \text{TradeAgreement}(\bta) \right] \left[W_E(\bta) + W_E^*(\bta) \right]
  \label{eq:jv}
\end{multline}

I model the executives' choice via the Nash bargaining solution where the disagreement point is the executives' welfare resulting from the Nash equilibrium in the non-cooperative game (i.e. in the absence of a trade agreement) between the legislatures.

The home executive's welfare is specified as follows:
\[
  W_E = \mathit{CS}_X(\tau) + \ga_E \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
\]
Note that this is identical to the welfare function for the legislature aside from the weight $\ga_E$ the executive places on the profits of the import industry, which is not a function of lobbying effort. \textbf{This setup does not strictly require that the executives are not lobbied, only that the extent to which the executives favor this industry cannot be \textit{directly} influenced by lobbying effort. The assumption is innocuous as long as the executives' preferences are not directly altered in a significant way by lobbying over trade.}

\textbf{I make one more assumption on $\ga(e,\ta)$:}

@@

\textbf{That is, even for the least favorable outcome of the lobbying process, the legislature will be at least slightly more protectionist than the executive regardless of the lobby's effort choice. This assumption is not essential, but it simplifies the analysis considerably, ensuring that trade agreement tariffs are less than trade war tariffs. It matches the empirical findings that politicians with larger constituencies are less sensitive to special interests and seems to represent particularly well the post-war U.S. where Congress has consistently been significantly more protectionist than the President.}\footnote{cfr. Lohmann and O'Halloran (1994) for the general argument, Destler (2005) on U.S. trade politics, and Grossman and Helpman (2005) for a formal model.}

\subsection{Information and Equilibrium Selection}
\label{sec:info}
I examine a simple class of equilibria that have the following features.

First, on information. Although uncertainty is present, information about it is symmetric. However, in line with the literature, I assume the lobby's contribution is not observable to the foreign legislature. The implication is that the lobby can directly influence only the home legislature, and so the influence of one country's lobby on the other country's legislature occurs only through the tariffs selected.\footnote{cfr. Grossman and Helpman (1995b), page 685.} Given the game's structure and the assumption that lobbying effort is not observable across international borders, players in the same country can take advantage of more information than those who are in different countries. That is, the solution concept employed here is perfect public equilibrium (PPE).

Second, whenever there is a possibility of multiple equilibria, I focus on the one that maximizes the welfare of the executives. Given the assumption that the executives are social welfare maximizers, this selection criterion puts results in terms of the maximal level of trade policy cooperation that is possible. It also allows the question of whether governments use trade agreements as political commitment devices to be answered in a straightforward way.

Finally, I assume that an external authority can ensure enforcement of the agreement in the one-shot game analyzed here.\footnote{Buzard (2015) considers a similar model in a repeated-game context and can therefore remove this restriction.}

>>>>

Figure~\ref{fig:ext} illustrates the timing of the game from the perspective of the home country.\footnote{This extensive form is for the case where the foreign legislature does not break the trade agreement.} First, the executives set trade policy cooperatively within the context of an international agreement by choosing the trade agreement tariffs $\bta = \left(\tau^a,\tau^{*a}\right)$.\footnote{I use the convention throughout of representing a vector of tariffs for both countries $(\tau,\tau^*)$ as a single bold $\bm{\tau}$.} Then the lobbies attempt to persuade the legislators in their respective countries to break the trade agreement by choice of lobbying effort $e_b$, which can take on any non-negative value.

@@

\subsection{Preferences}
\label{sec:pref}
As the economy is fully separable and the economic and political structures are symmetric, I focus throughout on the home country and the $X$-sector. The details are analogous for $Y$ and foreign.\footnote{These are the relevant sectors for analysis because they are the sectors in which there are lobbies.} I specify the welfare function of the home legislature as
\begin{equation}
  W_{\mathit{ML}} = \mathit{CS}_X(\tau) + \ga(e,\ve) \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
  \label{eq:ml}
\end{equation}
where $\mathit{CS}$ is consumer surplus, $\mathit{PS}$ is producer surplus, $\ga(e,\ta)$ is the weight placed on producer surplus in the import-competing industry, and $\mathit{TR}$ is tariff revenue.\footnote{Labor income $l$ could also be included in both median legislator and executive welfare. I omit it because its inclusion alters none of the results and only serves to complicate the exposition.} I model the decisions of the legislature as being taken by a median legislator with the weight the median legislator places on import-competing industry profits affected by the level of lobbying effort $e$ and a random variable $\ta$. I make the following assumptions on $\ga(e,\ta)$:

@@

Assumption~\ref{as:ga_c} formalizes the intuition that the legislature favors the import-competing industry more the higher is its lobbying effort, but that there are diminishing returns to lobbying activity.\footnote{In Ethier (2012), the returns to lobbying decline as the level of protection increases.} It allows political uncertainty to be correlated with the level of $e$, ruling out only (i) that higher effort levels make lower $\ga$-weights more likely and (ii) that increasing lobbying effort changes the structure of uncertainty so that higher $\ga$-weights become more likely at an accelerating pace. Assumption~\ref{as:ga_ta} \textbf{simply provides for an intuitive labeling so that} larger realization of the uncertainty variable $\ta$ increase the value of the political economy weight.

Given its expectations and the legislature's preferences, the home lobby chooses its lobbying effort ($e_b$ to influence the break decision and $e_n$ to influence the trade war tariff) to maximize the welfare function:
\begin{equation}
  \expect \left[U_L \right] = \Pr\left[ \text{TradeWar}(\bta,e_b,e_n) \right] \left[ \pi(\tau^{\mathit{n}}) - e_n \right] + \Pr\left[ \text{TradeAgreement}(\bta,e_b,e_n) \right] \pi(\tau^a) - e_b
  \label{eq:lv}
\end{equation}
where $\pi(\cdot)$ is the current-period profit and $\tau^a \ (\tau^\mathit{n})$ is the home country's tariff on the import good under a trade agreement (war).

In the first stage, the executives choose the trade agreement tariffs. I model this choice via the Nash bargaining solution with a disagreement point of executive welfare in the non-cooperative game between the legislatures. In this symmetric environment, the executives maximize their expected joint payoffs:
\begin{multline}
  \expect \left[W_E(\bta) \right] + \expect \left[ W_E^{*}(\bta) \right] = \Pr\left[ \text{TradeWar}(\bta) \right] \left[W_E(\bm{\tau}^{\mathit{n}}) + W_E^*(\bm{\tau}^{\mathit{n}}) \right] + \\ \Pr\left[ \text{TradeAgreement}(\bta) \right] \left[W_E(\bta) + W_E^*(\bta) \right]
  \label{eq:jv}
\end{multline}

The home executive's welfare is specified as follows:
\[
  W_E = \mathit{CS}_X(\tau) + \ga_E \cdot \mathit{PS}_X(\tau) + \mathit{CS}_Y(\tau^*) + \mathit{PS}_Y(\tau^*) + \mathit{TR}(\tau)
\]
This is identical to the legislature's welfare function except the weight the executive places on the profits of the import industry is not a function of lobbying effort. \textbf{This setup does not require that the executives are not lobbied, only that the extent to which the executives favor this industry cannot be \textit{directly} influenced by lobbying effort. The assumption is innocuous as long as the executives' preferences are not directly altered in a significant way by lobbying over trade. I make one more related assumption on $\ga(e,\ta)$:}

@@

\textbf{That is, even for the least favorable outcome of the lobbying process, the legislature is more protectionist than the executive regardless of the lobby's effort choice. This is not essential but it simplifies the analysis, ensuring that trade agreement tariffs are smaller than trade war tariffs. It matches the empirical finding that politicians with larger constituencies are less sensitive to special interests and represents well the post-war U.S. where Congress has consistently been more protectionist than the President.}\footnote{cfr. Lohmann and O'Halloran (1994) for the general argument, Destler (2005) on U.S. trade politics, and Grossman and Helpman (2005) for a formal model.}

\subsection{Information and Equilibrium Selection}
\label{sec:info}
I examine a simple class of equilibria that have three key features. First, although uncertainty is present, information about it is symmetric. However, in line with the literature, I assume the lobby's contribution is not observable to the foreign legislature. This implies that the influence of one country's lobby on the other country's legislature occurs only through the tariffs selected.\footnote{cfr. Grossman and Helpman (1995b), page 685.} Since players in the same country can take advantage of more information than those who are in different countries, the solution concept employed here is perfect public equilibrium (PPE).

Second, whenever there is a possibility of multiple equilibria, I focus on the one that maximizes the executives' welfare. Since I assume the executives are social welfare maximizers, this selection criterion puts results in terms of the maximal level of trade policy cooperation that is possible. It also allows the question of whether governments use trade agreements as political commitment devices to be answered in a straightforward way.

Finally, I assume that an external authority can ensure enforcement of the agreement.\footnote{Buzard (2015) considers a similar model in a repeated-game context and can therefore remove this restriction.}
